# Mixed effects models

A mixed model, also known as a hierarchical or multilevel model, in general contains both random (subject or group level) and fixed (population level) effects.

## Restricted maximum likelihood

The standard way for fitting mixed effects models in R is **lme4** [@R-lme4], which estimates the parameters via restricted maximum likelihood optimization.
A canonical example dataset is `sleepstudy`, bundled with the package, which records reaction times each day for 18 subjects restricted to three hours of sleep per night.
The data are visualised in Figure \@ref(fig:sleepstudy).

```{r sleepstudy, fig.cap = 'Average reaction time per day (ms) for 18 subjects in a sleep deprivation study. Each line represents one subject', echo = FALSE}
data(sleepstudy, package = 'lme4')
library(ggplot2)
theme_set(theme_minimal())
ggplot(sleepstudy) +
  aes(Days, Reaction, group = Subject) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, 8, 2)) +
  labs(x = 'Days of restricted sleep',
       y = 'Reaction time (ms)')
```

Overall, we would probably expect reaction times to increase as the subjects become more sleep-deprived.
But some subjects surely will have had faster reaction times than others, even before the study began.
This is modelled using a _random intercept_ coefficient.
Meanwhile, we might assume each day of sleep deprivation results in the same increase in reaction time for every participant, in which case a common, fixed slope coefficient is appropriate.
However, if instead we thought the daily effect of sleep deprivation varied between subjects, we could fit _random slopes_ (as well as the random intercept).

In **lme4** syntax, the random intercept model is
```{r sleep-lme4, message = FALSE}
library(lme4)
rand_intercept <- lmer(Reaction ~ Days + (1 | Subject), sleepstudy)
```
which yields the following summary output.
```{r echo = FALSE}
summary(rand_intercept)
```

Adding a random slope, we obtain
```{r sleep-lme4-2, message = FALSE, echo = 1}
rand_slope <- lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
summary(rand_slope)
```

The correlation between the random effects is close to zero.

## Markov chain Monte Carlo

Researchers who prefer a Bayesian approach can take advantage of the **rstanarm** [@R-rstanarm] and **brms** [@R-brms] packages, which allow specifying analogous Bayesian models, fitted using Hamiltonian Monte Carlo (rather than restricted maximum likelihood) but using the familiar **lme4**-style formula syntax.

If using default priors, the code is nearly identical.
The package **rstanarm** (applied regression modelling via RStan) has a series of functions with the prefix `stan_` that are analogous to common frequentist methods, e.g. `stan_glm` and, in this case, `stan_lmer`:
```{r arm-slope, results = 'hide', echo = 2:3, message = FALSE, cache = TRUE}
options(mc.cores = parallel::detectCores())
library(rstanarm)
arm_slope <- stan_lmer(Reaction ~ Days + (Days | Subject), sleepstudy)
```

```{r echo = FALSE, dependson='arm-slope'}
arm_slope
```

Meanwhile, **brms** (Bayesian regression modelling with Stan), which is designed for fitting Bayesian multilevel models, has a single workhorse function, `brm`, which also uses **lme4**-style formulae:

```{r brm-slope, results = 'hide', message = FALSE, cache = TRUE}
library(brms)
brm_slope <- brm(Reaction ~ Days + (Days | Subject), sleepstudy)
```

```{r echo = FALSE, depenson='brm-slope'}
brm_slope
```

As with other Stan wrapper packages, if you don't specify any prior distributions, then uninformative ones will be used by default.
This is usually fine in this case but you'll need to set explicit priors when fitting more complicated models.

All three of **lme4**, **rstanarm** and **brms** yield similar parameter estimates for the sleep deprivation study model with random intercepts and slopes:

- fixed intercept about 251 ms
- fixed slope about 10.5 ms per day
- random intercept standard deviation about 24 ms
- random slope standard deviation about 6 ms per day
- random intercept--slope correlation about 0.07--0.08

